A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-02-09 analysis/paper/paper.Rmd
edit based on BM's comments

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!--study social changes in pericolonial context using burial data as a proxy-->
Social changes in Indigenous societies when faced with colonial powers are commonly observed in many parts of the world, especially for European colonies where Indigenous economic, culture, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism or involvement in long-distance trade could also have impacts on Indigenous societies in a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further imply social complexity or social inequality where foreign goods from long-distance trade were often used by Indigenous people to express status [@Carter2015; @Dolfini2019]. Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. Recent development of network modeling provides new tools to explore the structure of the material configuration in archaeological data and dependencies between variables to understand past relationships, interactions, and underlying mechanisms [@Brughmans2013; @Brughmans2018; @Crabtree2015; @Erik2015; @Freeman2004; @Salvini2010; @Ostborn2015].

@@

<!---network analysis and the basic concepts-->
Different from a static characterization of burial variables, network modeling approaches have the potential to study social changes in a dynamic way by examining the intercorrelations between burials. Drawing largely on network science and graph theory, a key assumption in the interpretation of archaeological networks is that a phenomenon can be conceptualized as a network through abstraction and representation [@Brandes2013; @Collar2015; @Mills2017; @Peeples2019]. A network is viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. Among the network modeling methods, exponential random graph models (ERGMs) are one important family of statistical graphs to explore complex networks [@Ghafouri2020]. However, the computational difficulties and the sensitivity to uncertainties to some extent limits its applications [@Caimo2014]. In this paper, we try to go beyond ERGMs with a novel Bayesian approach to investigate the formation of material connections between burials and changes in network structures by examining evidence from a case study in northeastern Taiwan.

<!--research aim and questions with archaeological specific case study-->
We study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. We argue that the involvement in the long-distance trade stimulated by the Europeans might lead to increased social inequality at Kiwulan based on the theoretical framework of corporate-network strategies [@Blanton1996; @Feinman2000]. We test our hypothesis that changes in power structure from a corporate mode to a network mode can be observed in the burial network after the European arrival in 17th century Taiwan. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming the unequal social positions represented that might hint at social heterogeneity influence by the European arrival? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

<!-- archaeological background in NE Taiwan in general -->
Archaeological sites in northeastern Taiwan show the evidence of accumulation patterns of imported goods, such as ornaments and porcelains, in residential areas and burial contexts around the same time as European contacts in the 17th century [@Chen2007; @LWandBM2020ornament]. Imported goods were treated as prestige or wealth in those Indigenous societies that were commonly mentioned in Spanish historical records in the 17th century (1624-1642) [@Borao2009; @LiandWu2006]. For example, a document written by a Spanish priest described a Spanish soldier using carnelian beads to exchange for natural resources with local Indigenous people, because of their high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as more influential or having higher status in their community [@LiandWu2006]. Despite the fact that we recorded an uneven distribution of imported goods in burials, the degree of differentiation across burials over time remained unclear.

<!--specific case from Kiwulan-->
To explore indirect colonial impacts on Indigenous societies in northwestern Taiwan, our case study focuses on Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age settlement (1350-1850 AD) [@Chen2007]. Excavations revealed abundant trade goods, including imported ceramics and stonewares, metal artifacts, glass beads, and stone beads, indicative of frequent trade activities. The introduction of foreign prestige goods to local Indigenous communities might have occurred before European presence became established, but was maintained and amplified after European arrival due to local competition for power [@Borao2009; @LiandWu2006; @LWandBM2020ornament; @Ueda2016]. A general model to summarize this situation in northeastern Taiwan might assume that the influence of a colonial power, combined with high local values attached to imported goods, might lead to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution in Northeastern Taiwan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

@@

## Exponential random graph models in a Bayesian framework  {-}

<!-- ERGM in archaeology, and limitations --> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively. However, those studies also point out some limitations of ERGMs, such as sensitivity to missing data and less able to handle uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed "doubly intractable" since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. The typical criteria for interpreting a significant effect of a parameter is that its credible interval does not contain zero [@Harris2013, pp. 51]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data is missing.

# Materials {-}

@@

<!-- archaeological setting -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) (Figure \@ref(fig:figure-KWL-map)) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until Han Chinese immigrants arrival). The discord over the discussion of Kiwulan social organization could be associated with chronological differences that have not yet been well studied.

@@

<!-- chronology -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [@LWandBM2020ornament; @Chen2007]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. There are `r disturb` burials heavily disturbed by modern construction that are also excluded because we cannot determine their chronology. Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. Previous studies indicate that these items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. Historical accounts written by Europeans mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006].

@@

## Hypothesis and construction of networks {-}

<!-- our hypothesis --> 
Our "pericolonial impact hypothesis" is that rare, high quality burial goods accumulated in only a few individuals' burials after the European presence because of increased social inequality at Kiwulan due to differential access to these goods. A related effect of the European presence in northeastern Taiwan may have been manipulation of the European colonial image by ambitious local Indigenous individuals for building their personal status and power, recognized as network strategies to achieve power [@Drennan2010; @Feinman2000]. If social inequality gradually increased in Kiwulan as we hypothesize, then we expect to observe a network with higher degree (popularity) and less transitivity (cohesion), along with greater differences in wealth accumulation according to burial goods. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of goods between burials. Since trade beads are commonly found across burials with substantial differences in quantities, we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting based on historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three ranks, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculated by the total number of prestige items from burial context over the number of prestige items in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

```{r figure-network-diagrams, fig.cap = "A: Burial network before the European arrival, B: Burial network after the European arrival. The size of each node is proportional to 'node degrees' or the number of connections to a node. The thickness of tie represents the number of goods in common between two nodes."}
knitr::include_graphics(here::here("analysis", "figures", "005-network-diagrams.png"))
```

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], where model specification and simulation procedures is based on statnet package [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for the model specification. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the tendency of tie forming. For example, age-homophily means people of the same age tend to have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to understand the preferential tendency of formation of relationships. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
knitr::kable(ergm_terms,
             caption = "Network parameters in ERGMs used for model specifications with associated interpretation for burial relations.")
```

@@

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter = factor(parameter,
                            levels=c("edges", "nodematch.age", 
                                     "nodematch.sex", "nodematch.ritual",
                                     "nodematch.value", "gwesp",	
                                     "gwdeg", "dyadcov.distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.075), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.3) %>% 
  width(j = 2:9, width = 0.65) %>% 
  set_caption("Estimated posterior means, medians, and 95% credible intervals for each network parameter of two models.")
ft
```

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns to infer interactions and relationships between entities. Statistical approaches to network analysis, such as exponential random graph models (ERGMs), provide a way to test hypotheses on dynamic processes of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification of uncertainty, and robust model evaluation of network properties. We tested a hypothesis of social change relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with increased social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!---network analysis and the basic concepts-->
Network analysis has been increasingly used by archaeologists to explore rational data to understand past relationships, interactions, or overall structures of observed phenomena [@Brandes2013; @Mills2017; @Peeples2019]. A network can be viewed as a patterned aggregation that includes individual elements, pair-wise relationships, and an overall structure showing global patterns represented in the data. Recently, network modeling provides many new tools to examine statistically the intercorrelations between individual elements in a dynamic way. Among network modeling methods, exponential random graph models (ERGMs) are one important family of statistical graphs to explore complex networks [@Harris2013; @Ghafouri2020]. However, the computational difficulties and the sensitivity to uncertainties to some extent limits its practical applications [@Caimo2014]. In this paper, we go beyond ERGMs with a novel Bayesian approach to investigate the formation of material connections using burial data from a case study in Iron Age northeastern Taiwan to understand social changes during foreign contact.

<!--study social changes in pericolonial context using burial data as a proxy-->
Social changes in Indigenous societies, when faced with colonial powers, are commonly observed in many parts of the world, especially European colonies where Indigenous economic, cultural, or socio-political aspects were substantially impacted [@Dietler2005; @Silliman2005; @Voss2005]. Recent studies demonstrate that the indirect effect of colonialism, or involvement in long-distance trade, could also have impacts on Indigenous societies, this is known as a pericolonial context [@Acabado2017; @Trabert2017; @LWandBM2020ornament]. To explore social changes, burials are important evidence to understand past societies because material culture and biological records of burial behaviors can represent the social ranking or identities of the deceased, and social relations between them [@Binford1971; @Drennan2010; @Saxe1970]. Burial treatments can further indicate social complexity or inequality where foreign goods from long-distance trade were used by Indigenous people to express status [@Carter2015; @Dolfini2019]. Usually, burial studies are based on characterizing each physical trace or burial variable, such as biological records, grave forms and goods, or ritual behaviors, and finally combining those individual observations to infer past societies [@Byrd1995; @Seikel2011]. Recent development of network modeling provides new tools to explore the structure of the material configurations and dependencies between variables to understand past relationships, interactions, and underlying mechanisms [@Brughmans2013; @Brughmans2018; @Crabtree2015; @Erik2015; @Freeman2004; @Salvini2010; @Ostborn2015].

@@

# Exponential random graph models in a Bayesian framework  {-}

<!-- ERGMs in archaeology--> 
Exponential random graph models (ERGMs) first appeared in archaeology with @Brughmans2014, who studied Iron age settlement patterns in Southern Spain by modeling inter-settlement visibility networks and visual control at 159 sites. By fitting models with the observed network of archaeological data, they proposed that ERGMs is a promising method for exploration and hypothesis testing for social processes. Similarly, @Amati2019 modeled three networks consisting of 15 sites (AD 100 to 400) in the Caribbean to explore interaction mechanisms, including proximity, inter-cultural items, and pottery types. By comparing hypothesized networks with the observed sites, they found that the presence of hub sites can be efficiently explained by multiple interdependent mechanisms instead of only one variable exclusively.

<!--limitations of ERGMs-->
However, those studies also point out some limitations of ERGMs, such as sensitivity to missing data and less able to handle uncertainty. Also, it is difficult in ERGMs to estimate model parameters and interpret the result due to intractable likelihood normalizing constants and model degeneracy [@Caimo2014; @Jin2013]. A normalizing constant is a function of the model parameter for making probability distributions integrate to one, which becomes harder to compute with a larger set of networks [@Caimo2017bayesian]. This is also termed “doubly intractable” since both the likelihood normalizing constant and the marginal likelihood (the evidence of the posterior) are hard to derive [@Caimo2013; @Lyne2015]. Model degeneracy is another issue where probability models tend to overestimate a small number of extreme graphs by assigning too much weight, e.g., in extreme cases such as empty (all nodes unconnected) or complete graphs (all nodes connected) [@Schweinberger2011; @Caimo2014]. One solution to these limitations is available by implementing ERGMs by means of a Bayesian framework.

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to ERGMs are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. With the exchange algorithm, Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives better convergence results. This enables us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. By fitting an ERMG with the approximate exchange algorithm, a Bayesian approach generates posterior probabilities that incorporate our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by examining the posterior mean and 95% credible intervals [@Caimo2017bayesian]. In addition, Bayesian approaches are useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on average, 80% of the ties when a third of data is missing.

@@

<!--archaeological specific case study and our model-->
We use a Bayesian ERGM to study social changes in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan (Figure \@ref(fig:figure-KWL-map)), which was occupied from the 14th to 19th centuries. This occupation period includes the time before the European arrival, the presence of the Spanish and the Dutch in the 17th century, and finally a large wave of Han Chinese in the 19th century [@Chen2007]. Our general model proposes that the influence of a colonial power, combined with high local values attached to imported goods, leads to increased social inequality due to competition among individuals [@Brumfiel1994; @Clark1994]. The observed uneven distribution of burial goods at Kiwulan may be explained as a result of unequal access to trade goods when Indigenous societies became more involved in the complex trade network stimulated by Europeans.

<!--theoretical framework and research questions-->
Using the theoretical framework of corporate-network strategies [@Blanton1996; @Feinman2000], we test a hypothesis that changes from a corporate mode (…) to a network mode (…) can be observed in the Kiwulan burial network after European arrival. We ask: (1) did European colonial activities result in increased social inequality in Indigenous society in ways that can be detected by analysis of burial networks? (2) if so, what are the major variables affecting or forming unequal social positions that might hint at social heterogeneity? By answering these questions, this case study helps to expand our understanding of European colonial effects on Indigenous groups in a pericolonial context.

@@

## Materials {-}

@@

<!-- Kiwulan burials and a review of current studies of social organization -->
We analyzed burial data collected from the excavation reports, and the original fieldwork notes for the upper component of Kiwulan (1350-1850 AD) [@Chen2007]. A total of 90 burials were unearthed from adjacent squares that provide continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, which is indicated by post-holes and _in-situ_ wooden posts, suggesting a well-organized spatial arrangement of houses. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on the comparative analysis of the frequencies of all burial goods. She found that the burials with rare prestige goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing the phases before and after European arrival. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period (before European colonization) to the post-European period (during and after European colonization until arrival of Han Chinese immigrants).

@@

<!-- Our chronology and principles of network building -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [for more details see @LWandBM2020ornament]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. We also excluded `r disturb` burials that were heavily disturbed by modern construction because we cannot determine their chronology. Based on the assumption that social status and associated relations can be represented by sharing similar prestige goods [cf. @Coward2013, pp. 252], we built networks where burials (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include gold-foil beads, carnelian beads, glass beads, Chinese porcelains, stonewares, gold foils, and fish-shaped ornaments. These items are considered as high-value across different archaeological contexts based on their rarity and descriptions in historical records [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. European historical accounts mentioned that those items were treated as prestige goods in Indigenous culture [@Borao2009; @LiandWu2006]. For example, a Spanish priest described a Spanish soldier exchanging carnelian beads for natural resources with local Indigenous people, because of the beads’ high value in Indigenous culture [@LiandWu2006]. In addition, Spanish visitors observed that an Indigenous person possessing more imported goods may have been recognized as having higher status in their community [@LiandWu2006].

@@

## Hypothesis and construction of networks {-}
<!-- our expectation based on hypothesis --> 
Our pericolonial impact hypothesis is that differential access to foreign prestige goods after the European presence led to an increased social inequality at Kiwulan, where social structure changed from a more corporate strategy to a more network strategy [@Blanton1996; @Feinman2000]. This would be reflected by the pattern of intercorrelations of foreign prestige goods in burial networks. If social inequality gradually increased as we hypothesize, then we expect to observe a network structure with higher prominence (popularity) and less transitivity (cohesion), along with associations with wealth accumulation according to burial goods. To test this hypothesis, we use Bayesian ERGMs to model the formation of network ties and the underlying mechanisms that shape relationships between people at Kiwulan. By comparing networks from the pre-European period and the post-European period we can examine the effects of foreign contact on community relationships at Kiwulan.

@@

<!-- burial data and the definition of tie formation  --> 
Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported ceramics, gold foils, and fish-shaped ornaments, we linked two burials when they both possess each type of goods (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and a burial value index. Ritual pottery was identified as locally made ceramics placed above an individual. Historical records link these ceramics to funeral feasting [@citation]. We calculated our burial value by summing values of all types of prestige goods in each burial context. The value of each type of prestige good is calculated by taking the number of prestige items in all burials, and then dividing this by total number of prestige items in one burial context. We then assigned the burials into three ranks according to their burial values, high (>`r high_value`, top 10 percent), above average (`r ave_value`-`r high_value`), and below average (<`r ave_value`), as an index of wealth. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary (the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means a mutual relationship where the tie between any two actors is bidirectional. The networks constructed based on these principles show that the network after the European presence has more node connectivity in general with some nodes having a larger number of connections (Figure \@ref(fig:figure-network-diagrams)).

```{r figure-network-diagrams, fig.cap = "A: Burial network before the European arrival, B: Burial network after the European arrival. The size of each node is proportional to node degrees or the number of connections to a node. The thickness of tie represents the number of goods in common between two nodes."}
knitr::include_graphics(here::here("analysis", "figures", "005-network-diagrams.png"))
```

@@

<!-- Network construction and specification -->
We quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014]. Table \@ref(tab:table-ergm-terms) lists the network parameters we used for the model specification [@Morris2008]. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality to be represented by endogenous network effects: low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to explore the importance of these variables in tie forming. For example, if age-homophily is important here, then people of the same age should have the same burial goods. We also include the physical distance between burials as an indicator of a kinship-based relations since the deceased from the same family were buried nearby [@LiandWu2006]. Our model may reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

```{r table-ergm-terms}
ergm_terms <- readxl::read_excel(here::here("analysis", "data", "raw_data", "ergm_term.xlsx")) 
knitr::kable(ergm_terms,
             caption = "Network parameters in ERGMs used for model specifications with associated archaeological interpretation for burial relations.")
```

@@

```{r table-posterior-stats}
# tidy data for the table of posterior stats 
posterior_esti_two_phases<-
  bergm_two_phases %>%
  pivot_longer(!phase, 
               names_to = "parameter", 
               values_to = "posterior") %>%
  mutate(parameter = factor(parameter,
                            levels=c("edges", "nodematch.age", 
                                     "nodematch.sex", "nodematch.ritual",
                                     "nodematch.value", "gwesp",	
                                     "gwdeg", "dyadcov.distance"))) %>% 
  group_by(parameter, phase) %>% 
  summarize(mean = round(mean(posterior), 2),
            median = round(quantile(posterior, 0.5), 2),
            `2.5%` = round(quantile(posterior, 0.025), 2),
            `97.5%` = round(quantile(posterior, 0.975), 2)) %>% 
  pivot_wider(names_from = phase, 
              values_from = c(mean, median, `2.5%`, `97.5%`)) %>% 
  select(parameter, ends_with(c("pre", "post")))

library(flextable)
# add column labels
ft <- flextable(posterior_esti_two_phases) %>% 
  set_header_labels(mean_pre = "pre-European", median_pre = "pre-European", 
                    `2.5%_pre` = "pre-European", `97.5%_pre` = "pre-European",
                    mean_post = "post-European", median_post = "post-European", 
                    `2.5%_post` = "post-European", `97.5%_post` = "post-European") %>%
  merge_at(i = 1, j = 2:5, part = "header") %>% 
  merge_at(i = 1, j = 6:9, part = "header") %>% 
  add_header_row(values = c("", "mean", "median", "2.5%", "97.5%",
                            "mean", "median", "2.5%", "97.5%"), top = FALSE ) %>% 
  align(j = 2:9, align = "right", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  width(j = ~ parameter, width = 1.3) %>% 
  width(j = 2:9, width = 0.65) %>% 
  set_caption("Estimated posterior means, medians, and 95% confidence intervals for each network parameter of two models. Confidence intervals that do not include zero idicate significant effects of parameters (Caimo, 2017).")
ft
```
